---
title: "Bayesian Hierarchical Modeling of Election Results"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

**Objective**: Smooth the daily estimations of the probability of a Republican win in each state, accounting for the hierarchical structure of the data, the effect of `Date`, and comparing different `Voice`s.

**Data Structure**:

- **Date**: Observation day.
- **Voice**: One of Direct, Fox, MSNBC, or BBC.
- **Trial**: Daily trials within each voice (90-100 per day).
- **State**: 50 states plus DC (51 in total).
- **Result**: Binary outcome (1 = Republican win, 0 = Democrat win).

# Table of Contents

1. [Install and Load Necessary Packages](#install-and-load-necessary-packages)
2. [Read and Prepare the Data](#read-and-prepare-the-data)
3. [Data Exploration and Preparation](#data-exploration-and-preparation)
4. [Specify the Bayesian Hierarchical Model](#specify-the-bayesian-hierarchical-model)
5. [Fit the Model Using `brms`](#fit-the-model-using-brms)
6. [Check Model Diagnostics and Address Convergence Issues](#check-model-diagnostics-and-address-convergence-issues)
7. [Extract and Visualize Smoothed Estimates](#extract-and-visualize-smoothed-estimates)
8. [Compare Different Voices](#compare-different-voices)
9. [Summary and Next Steps](#summary-and-next-steps)
10. [References](#references)

# 1. Install and Load Necessary Packages

```{r}
# Install packages if not already installed
if (!require("brms")) install.packages("brms")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("bayesplot")) install.packages("bayesplot")

# Load the libraries
library(brms)
library(tidyverse)
library(bayesplot)
```

# 2. Read and Prepare the Data

```{r}
# Read your CSV file (replace with your actual file path)
data <- read.csv('/Users/jaredblack/GitHub/ElectionGPT/data/panel_election_results_state.csv', stringsAsFactors = FALSE)

# Convert 'Date' to Date type using the correct format
# The dates are in "8/13/24", which corresponds to "%m/%d/%y"
data$Date <- as.Date(data$Date, format = "%m/%d/%y")

# Check for NAs after conversion
if (any(is.na(data$Date))) {
  warning("There are missing dates after conversion. Please check the date format.")
}

# Standardize the 'Voice' column
data$Voice <- tolower(data$Voice)
data$Voice <- tools::toTitleCase(data$Voice)

# Ensure 'Voice', 'Trial', and 'State' are factors
data$Voice <- as.factor(data$Voice)
data$Trial <- as.factor(data$Trial)
data$State <- as.factor(data$State)
```

# 3. Data Exploration and Preparation

## 3.1. Aggregate the Data

We need to aggregate the binary results to get counts of successes (Republican wins) and total trials per state per date per voice.

```{r}
# Summarize data by Date, State, and Voice
summary_data <- data %>%
  group_by(Date, State, Voice) %>%
  summarise(
    successes = sum(Result),  # Number of Republican wins
    trials = n(),             # Total number of trials
    .groups = 'drop'          # Ungroup after summarizing
  )
```

## 3.2. Prepare Predictors

### Center and Scale `Date`

Centering and scaling continuous predictors can help with model convergence.

```{r}
# Convert Date to numeric (number of days since first date)
summary_data$Date_num <- as.numeric(summary_data$Date - min(summary_data$Date))

# Center and scale Date_num
summary_data$Date_num_c <- scale(summary_data$Date_num, center = TRUE, scale = TRUE)
```

### Check for Perfect Separation

Perfect separation can cause convergence issues.

```{r}
# Check for perfect separation
separation_check <- summary_data %>%
  mutate(proportion = successes / trials) %>%
  filter(proportion == 0 | proportion == 1)

# Print problematic observations
print(separation_check)
```

**Action**: If you find observations with perfect separation, consider adding informative priors or excluding them (with justification).

# 4. Specify the Bayesian Hierarchical Model

Include `Date` and `Voice` as predictors, with random effects for `State` and random slopes for `Date`.

```{r}
# Define the model formula including interactions
model_formula <- bf(
  successes | trials(trials) ~ Date_num_c * Voice + (Date_num_c | State)
)
```

# 5. Fit the Model Using `brms`

Adjust model settings to address convergence issues.

```{r}
# Define priors
priors <- c(
  prior(normal(0, 1), class = "Intercept"),          # Prior for the intercept
  prior(normal(0, 0.5), class = "b"),                # Priors for coefficients
  prior(student_t(3, 0, 1), class = "sd"),           # Prior for random effects SD
  prior(lkj(2), class = "cor")                       # Prior for correlations
)

# Fit the model
fit <- brm(
  formula = model_formula,
  data = summary_data,
  family = binomial(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 10000,          # Increased number of iterations
  warmup = 5000,         # Increased warm-up period
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 20)  # Increased adapt_delta and max_treedepth
)
```

**Note**: Be prepared for longer computation times due to increased iterations and treedepth.

# 6. Check Model Diagnostics and Address Convergence Issues

```{r}
# Summarize the fitted model
summary(fit)

# Plot trace plots for diagnostics
plot(fit)

# Posterior predictive checks
pp_check(fit)
```

**If Warnings Persist**:

- Examine the `Rhat` values (should be close to 1).
- Check effective sample sizes (`Bulk_ESS` and `Tail_ESS`).
- If issues remain, consider simplifying the model or adjusting priors further.

# 7. Extract and Visualize Smoothed Estimates

## 7.1. Prepare New Data for Predictions

Ensure that `new_data` includes all variables used in the model (`Date_num_c`, `Voice`, `State`, and `trials`).

```{r}
# Create new data for predictions
new_data <- summary_data %>%
  select(Date, Date_num_c, State, Voice, trials) %>%
  distinct()
```

## 7.2. Generate Predicted Probabilities

```{r}
# Generate predicted probabilities
predictions <- fitted(fit, newdata = new_data, re_formula = NULL, summary = TRUE)

# Combine predictions with new_data
predictions_df <- cbind(new_data, predictions)
```

## 7.3. Plot the Results

Plot the smoothed probabilities over time for a specific state, comparing different voices.

```{r}
# Choose a state to plot
state_to_plot <- "California"  # Replace with desired state

# Filter data for the selected state
plot_data <- predictions_df %>% filter(State == state_to_plot)

# Plot the smoothed probabilities
ggplot(plot_data, aes(x = Date, y = Estimate, color = Voice)) +
  geom_line() +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5, fill = Voice), alpha = 0.2) +
  labs(
    title = paste("Smoothed Probability of Republican Win in", state_to_plot),
    x = "Date",
    y = "Estimated Probability"
  ) +
  theme_minimal()
```

# 8. Compare Different Voices

By including `Voice` as a predictor and fitting the combined model, you can directly compare the effects of different voices.

## 8.1. Extract Coefficient Estimates

```{r}
# Extract fixed effects
fixef(fit)
```

## 8.2. Visualize the Effects

Plot the estimated coefficients for `Voice` and `Date_num_c`.

```{r}
# Coefficient plot
mcmc_plot(fit, type = "areas", pars = c("b_Intercept", "b_Date_num_c", "b_VoiceFox", "b_VoiceMsnbc", "b_VoiceBbc"))
```

## 8.3. Posterior Predictive Checks by Voice

```{r}
# Posterior predictive checks by Voice
pp_check(fit, group = "Voice")
```

# 9. Summary and Next Steps

- **Model Diagnostics**: Ensure convergence by checking `Rhat`, `ESS`, and trace plots.
- **Address Warnings**: If warnings about treedepth or ESS persist, consider further model adjustments.
- **Interpret Results**: Analyze the estimated effects of `Date` and `Voice` on the probability of a Republican win.
- **Visualization**: Create comparative plots for different states and voices to visualize trends.
- **Reporting**: Document your methodology, results, and any challenges encountered.

# 10. References

- **`brms` Package Documentation**: [CRAN - brms](https://cran.r-project.org/package=brms)
- **Stan Modeling Language**: [Stan User's Guide](https://mc-stan.org/users/documentation/)
- **Bayesian Data Analysis**: Gelman et al.
- **Visualization**: [ggplot2 Documentation](https://ggplot2.tidyverse.org/)
- **Bayesplot Package**: [Bayesplot Documentation](https://mc-stan.org/bayesplot/)

---

**Additional Tips**

- **Computational Considerations**: Long models may require substantial computational resources. Monitor your system's performance.
- **Saving Models**: Use `saveRDS(fit, "fit_model.rds")` to save your fitted model and avoid re-running lengthy computations.
- **Further Analysis**: Consider exploring interaction effects or adding more predictors if appropriate.
- **Collaboration**: Discuss your modeling approach with peers or advisors for additional insights.

---

**Let me know** if you have any questions or need further assistance with specific parts of the code or interpretation of the results. Good luck with your analysis!
```

---

**Notes on the Updates**:

- **Included `Date` as a Predictor**: `Date_num_c` (centered and scaled `Date`) is included in the model to capture temporal trends.
- **Comparison of `Voice`s**: `Voice` is included as a predictor, and interactions between `Date_num_c` and `Voice` are modeled to compare different voices.
- **Fixed Errors**:
  - **`trials` Variable**: Included `trials` in `new_data` for predictions.
  - **`fit_rs` Object**: Since we're now using `fit` as the combined model with `Date` and `Voice`, `fit_rs` is no longer needed.
- **Adjusted Model Settings**:
  - Increased `max_treedepth` to 20.
  - Increased `iter` to 10000 and `warmup` to 5000.
  - Adjusted priors to be more informative, which can help with convergence.
- **Centered and Scaled Predictors**: Centering and scaling of `Date_num_c` can improve sampling efficiency.

**Remember to**:

- Replace `'path/to/your/data.csv'` with the actual path to your data file.
- Ensure that you have sufficient computational resources to run the model with increased iterations and `max_treedepth`.
- Monitor model diagnostics and adjust as necessary.

**If Warnings Persist**:

- Consider simplifying the model by removing random slopes or interactions.
- Experiment with different priors or modeling strategies.
- Check for data issues such as perfect separation or sparse data in some groups.

---

**Feel free to reach out** if you have any questions or need further assistance. I'm here to help you get your analysis running smoothly!